The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-04-24 Paper/jaccardbaetica.tex
new changes xavi

%This paper explore--- resumir en dos líneas el objeto de estudio y definir que ha sido usado para ello dispersion of culture
\xavi{este aim es demasiado general}
This paper aims to understand the production dynamics in connection with different areas within the Roman Empire. Specifically, our work pretends to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres. To do this, an population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

@@

The high demand of Roman provinces stimulated by the good condition of the \textit{Baetica} lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially for almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were made to support this high demand of the Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

@@

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of \textit{Baetica} province in the Roman organisation. Indeed, archaeological evidence shows a highly specialised production with a long activity in this specific area with apparently few changes \citep{remesal_anforas_2004}.

%hipotesis: si esto es así entonces es asá

The assumption of this study is analysing the effect of the production patterns between different centres. We are specially interested in identifying links between producers and consumption centres using amphoric stamps. To do that, we use the CEIPAC database to collect the stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.

Two case studies have been studied in order to analyse the relation between production centres (\textit{Baetica}) and consumption centres (\textit{Britannia} and \textit{Germania}). In the case of \textit{Baetica} province, we want to identify the role of the stamps in the organisation of the workshop; in Roman provinces, our aim is detecting groups of stamps concentrated in an area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could be identified by different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamp in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province. 
%\xavi{affected? No sería "identified"? Es decir, la correlación es la evidencia, pero la causa sería la interacción económica...no?}

Therefore, this study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. 
%\xavi{no veo la parte evolutiva en el paper; por otra parte, si es relevante no la lanzaría aquí al vuelo; hace falta explicar por qué es necesaria, qué implica, etc.}

%this paper aims to highlight the production dynamic in connection with the proximity between amphora workshops and stamps.

\section{Material and methods}

Our case study examines the relationship between the distribution of amphoric stamps and the workshops. In particular, we study the distribution of amphoric stamp in order to identify a correlation between geographical distance and amphoric stamps. Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of stamps, b) stamps located in close workshops share similar traits and c) Low mobility of amphoric stamps to other regions: stamps always stay in the same region.

The amphora workshops were situated in different locations in \textit{Baetica} province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop}).

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production that it can be difficult for specifying an accurate chronology. This could be due to two reasons; firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible; secondly, Dressel 20 was produced during almost three centuries with apparently few changes \citep{berni_piero_chapter_2017}.

@@

@@

The frequency of stamps per workshop can be seen in Fig. \ref{stamps}. Most stamps are distributed in only one workshop whereas few workshops concentrate a high frequency of different amphoric stamps. The type of distribution is also frequent in the Roman economy where we observed a self-organised complex system pattern typical of a free market \citep{bayesian_2018}.

\xavi{quieres decir que esto es una power base? yo creo que no, no es tan bestia no? Yo lo matizaría}

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequencystamp.pdf}
\caption{Histogram with the frequency of the number of stamps per workshop. The distribution is widely diverse with most stamps concentrated in one workshop}
\label{stamps}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps for each area. Colors are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{frequency}
\end{figure}

@@

\subsection{Hieralchical clustering}

\maria{que pongo aquí}

>>>>

%\xavi{este aim es demasiado general}
%This paper aims to understand the production dynamics in connection with different areas within the Roman Empire. 
This paper aims to study the olive oil market connection between province by calculating the similarity index. Specifically, our work pretends to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

Two case studies have been studied in order to analyse the relation between production centres (\textit{Baetica}) and consumption centres (\textit{Britannia} and \textit{Germania}). In the case of \textit{Baetica} province, we want to identify the role of the stamps in the organisation of the workshop; in Roman provinces, our aim is detecting groups of stamps concentrated in an area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could be identified by different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamps in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province. 
%\xavi{affected? No sería "identified"? Es decir, la correlación es la evidencia, pero la causa sería la interacción económica...no?}

%Our case study examines the relationship between the distribution of amphoric stamps and the workshops.
In particular, we study the distribution of amphoric stamp in order to identify a correlation between geographical distance and amphoric stamps. Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of stamps, b) stamps located in close workshops share similar traits and c) Low mobility of amphoric stamps to other regions: stamps always stay in the same region.

To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

@@

The high demand of Roman provinces stimulated by the good condition of the \textit{Baetica} lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially for almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were made to support this high demand of the Roman Empire. The amphora workshops were situated in different locations in \textit{Baetica} province, along the rivers Guadalquivir and Genil (see Fig.\ref{workshop}).

The location of the workshops with a riverine connectivity allowed to set up a connection with the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}.

@@

Some stamps show a more specific chronology while the majority of them display a large activity of production that it can be difficult for specifying an accurate chronology. This could be due to two reasons; firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible; secondly, Dressel 20 was produced during almost three centuries with apparently few changes \citep{berni_piero_chapter_2017}.

@@

Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of \textit{Baetica} province in the Roman organisation. Indeed, archaeological evidence shows a highly specialised production with a long activity in this specific area with apparently few changes \citep{remesal_anforas_2004}.

@@

The assumption of this study is analysing the effect of the production patterns between different centres. We are specially interested in identifying links between producers and consumption centres using amphoric stamps. To do that, we use the CEIPAC database to collect the stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}. Therefore, this study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. 
%\xavi{no veo la parte evolutiva en el paper; por otra parte, si es relevante no la lanzaría aquí al vuelo; hace falta explicar por qué es necesaria, qué implica, etc.}

@@

The frequency of stamps per workshop can be seen in Fig. \ref{stamps}. Most stamps are distributed in only one workshop whereas few workshops concentrate a high frequency of different amphoric stamps. The type of distribution is also frequent in amphora production where we observed a self-organized complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}.

%The type of distribution is also frequent in the Roman economy where we observed a self-organised complex system pattern typical of a free market 
%\xavi{quieres decir que esto es una power base? yo creo que no, no es tan bestia no? Yo lo matizaría}

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequencystamp.pdf}
\caption{Histogram with the number of stamps (X axis) and the frequency of distribution per workshop (Y axis). The distribution is widely diverse with most workshops concentrated one stamp}
\label{stamps}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps (X axis) for each \textit{conventus} (Y axis). Colors are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{frequency}
\end{figure}

@@

\subsection{Hierarchical clustering}

Morisita-Horn index was used to compute a matrix to group sites with the stamps codes that share similarities. The matrix was then visualised using hierarchical clustering. The algorithm was selected to cluster similar groups in order to analyse the relationship between groups of sites and the distribution of similar stamp codes.

\maria{mejorar}
